### 2023-03-21 products/submissions/SocSci/final_materials/main.tex
Prepare final manuscript for Soc Sci

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\KOMAoption{captions}{tableheading}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Differences in the Risk of Grade Retention for Biracial and Monoracial Students in the US, 2010-2019},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Differences in the Risk of Grade Retention for Biracial and
Monoracial Students in the US, 2010-2019\thanks{Supplementary materials
provided with this article include full model results upon which figures
are based as well as any sensitivity analysis described herein. All code
and data for this project are available at
\url{https://osf.io/4fevt/?view_only=4abc6d86595c4313a8d4792471e9bc0d}.}}
\author{}
\date{}

\begin{document}
\maketitle
\begin{abstract}
Understanding how outcomes for biracial individuals compare to their
monoracial peers is critical for understanding how patterns of racial
inequality in the contemporary United States might be shifting. Yet, we
know very little about the life chances of biracial individuals due to
limitations in most available data sources. In this article, I utilize
American Community Survey data from 2010-2019 to examine the risk of
being clearly behind expected grade among biracial and monoracial K-12
students, helping to fill a gap in our understanding. With large sample
sizes for most biracial groups, I am able to estimate grade retention
risk for biracial students with enough precision to differentiate even
modest differences in risk relative to monoracial groups. The results
indicate that for most biracial groups, biracial students have risk
similar to their monoracial constituent group with lower risk. Although
biracial students tend to have favorable family resource
characteristics, controlling for these characteristics does little to
change the overall placement of their outcomes.
\end{abstract}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, breakable, boxrule=0pt, frame hidden, sharp corners]}{\end{tcolorbox}}\fi

\setstretch{1.5}
\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

The experiences of mixed race individuals will play an important role in
the future of race in the US. A rise in interracial unions since the
1960s has resulted in a ``biracial baby boom'' of individuals whose
parents cross racial lines (Root 1992) and this demographic surge may
have transformative effects on racial boundaries (Alba 2020; Hochschild,
Weaver, and Burch 2012). While racial mixing is nothing new to US
history (Gullickson and Morning 2011; Morning 2000), this biracial boom
is notable for its size and has occurred in an environment in which
historical norms of racial classification may be eroding.

The transformative potential of this biracial baby boom is contingent on
both how these biracial individuals identify their own race and how they
are sorted into a prevailing system of racial inequality. While
extensive scholarship documents the identification and classification
decisions made by or on behalf of multiracial individuals (Bratter and
Heard 2009; Brunsma 2005; Davenport 2016; Herman 2004; Holloway et al.
2009; Liebler 2016; Qian 2004; Roth 2005; Xie and Goyette 1997), only a
smattering of research has documented how the life chances of biracial
individuals compare to their monoracial peers (Campbell 2009; Herman
2009; Kao 1999). In this article, I expand on this work by examining the
risk of grade retention for K-12 students whose parents identify with
different races.

By examining this outcome, I am able to overcome two important
limitations that have hampered prior work on the life chances of
biracial individuals. First, using self identification of biracial
respondents to analyze outcomes is problematic. Prior research indicates
that individuals whose parents belong to different races do not
consistently identify with both races (Doyle and Kao 2007; Gullickson
2019; Liebler et al. 2017). Furthermore, social status and social
mobility can predict racial identification generally (Saperstein and
Gullickson 2013; Saperstein and Penner 2012) and social status is
related to the identification choices of biracial individuals
specifically (Davenport 2016; Roth 2005). Thus, differences in outcomes
among biracial individuals may in part produce differences in
identification, rather than the other way around.

@@

Biracial children who reside with both biological parents can be
identified in large scale data sources such as the Census. Researchers
have previously utilized this feature to examine the classification
choices made on behalf of these children (Roth 2005; Xie and Goyette
1997). Analyzing outcomes for such children is more difficult, for the
simple reason that they do not have many measurable outcomes yet.
However, grade retention is measurable and important even for young
children. Being held back in school is a relatively common occurrence
that has consequences for later educational transitions (Jimerson 2001)
and is disproportionately applied by race (Warren, Hoffman, and Andrew
2014). Grade retention is one of the earliest stratifiers of life
chances that an individual will experience.

@@

Discrimination and bias by race play an important role in overall levels
of racial inequality (Pager and Shepherd 2008; Quillian 2006).
Discrimination need not be consciously recognized by actors, but may
instead reflect cognitive biases and schema that individuals unknowingly
rely upon and reinforce in social interactions (Quillian 2006).
Discrimination can also be systemic in nature when biases are built into
and reinforced by organizations and institutional systems (Reskin 2012).
In our current ``colorblind'' regime (Bonilla-Silva 2006), such systems
may be formally race neutral but produce disparate impacts by race
(Pager and Shepherd 2008).

@@

The strongest historical precedent for understanding such classification
in the US is the ``one-drop rule'' norm governing classification as
Black. According to the one-drop rule, individuals of known Black
ancestry are considered exclusively Black, regardless of other
ancestries (Davis 1991). Davis (1991) generalizes this practice into the
concept of \emph{hypodescent} in which individuals with mixed racial
ancestry will be classified as members of the lower status group. A
parallel pattern of \emph{hyperdescent} implies that individuals of
mixed racial ancestry will be classified as members of the higher status
group. This pattern of hyperdescent has historically governed the
experiences of individuals with American Indian ancestry (Snipp 1989;
Wolfe 2001).

Although the hypodescent/hyperdescent paradigm is often used to frame
the experiences of multiracial individuals, its generalizability and
contemporary relevance is questionable. Hypodescent and hyperdescent are
observed in two specific historical cases involving Black ancestry and
American Indian ancestry, respectively. As Iverson et al. (2022) note,
these two cases could just as parsimoniously be explained by a
``dominance'' model which ranks ancestries by their tendencies to be
supercessive or recessive in determining identification and
classification.

The development of particular norms may also reflect the historical
regime in which they were developed (Gullickson and Morning 2011;
Iverson et al. 2022). In this case, neither the hypodescent/hyperdescent
or dominance paradigm may help us to fully understand the experiences of
those of Latino and Asian ancestry due to more recent processes of
racialization. Furthermore, the historical norms for all groups may be
transforming in the post-Civil Rights era.

@@

Second, biracial people's experience of discrimination may be
``softened'' because a more ambiguous physical appearance provokes less
racial antipathy. The US already has a long history of this kind of
softening. Because of the one-drop rule, Black individuals vary
substantially in skin tone and those individuals with a lighter skin
tone have better outcomes (Hughes and Hertel 1990; Keith and Herring
1991; Monk 2014) and are viewed more favorably by both Blacks and Whites
(Maddox and Gray 2002). Thus, light skin moderates the stigma of
Blackness even among monoracial individuals.

Third, biracial individuals may be classified into an ambiguous
``middleman'' minority position. This kind of middle tier status for
individuals of mixed race is common in Latin America (Telles and Sue
2009) and Bonilla-Silva (2004) has argued that such a system may be
emerging in the contemporary US as well. In this case, many biracial
groups may occupy a nebulous and ill-defined middleman or buffer class
position in-between whites and darker-skinned non-whites, helping to
cement a new kind of racial hierarchy.

@@

Racial inequality is at least partly explained by different
distributions of material, cultural, and social resources across racial
groups (Conley 1999). On average, socioeconomic background
(e.g.~parental income, wealth, and education) differs by race and such
socioeconomic background is consequential for outcomes. Other less
materially defined resources also vary across groups such as cultural
styles, social networks, and neighborhood quality. These differences in
background resources reflect the accumulation of historical processes of
discrimination and, when combined with contemporary processes of
discrimination, magnify the overall level of racial inequality we
observe.

How do differences in the distribution of resources affect biracial
individuals? Naively, we might expect biracial individuals to inherit
parental resources roughly halfway between their two constituent groups,
on average (Chew, Eggebeen, and Uhlenberg 1989). Prior research looking
at biracial children in the Census data has found some evidence to
support this argument (Chew et al. 1989). In such a case, the resources
of biracial individuals would predict outcomes in-between the two
constituent groups, barring other mechanisms.

@@

Status exchange theory predicts that given an unequal racial hierarchy,
individuals from a lower status racial group will be positively selected
on socioeconomic characteristics into intermarriage with a higher status
group, while individuals from a higher status racial group will be
similarly negatively selected into the same marriage (Davis 1941; Fu
2001; Merton 1941). Prior research has shown substantial evidence of
this pattern in the case of Black male/White female marriages in the US,
but weaker evidence for other types of unions (Fu 2001; Gullickson 2006;
Hou and Myles 2011; Kalmijn 2010). Even in cases where status exchange
is present, its net effect on resources is complex because status
exchange positively selects one partner and negatively selects the other
partner. Therefore the two effects may somewhat cancel each other out in
terms of the overall distribution of resources across parents.

@@

The experience of hybridity itself may also positively or negatively
affect outcomes for biracial individuals. According to ``marginal man''
theory, biracial individuals will find it difficult to fit in due to the
dissonance of an identity that crosses strong and salient racial
boundaries (Park 1928; Stonequist 1935). This liminal identity conflict
will result in negative mental health experiences that could feed into
other negative outcomes as well. Evidence for such negative mental
health experiences, however, has been mixed (Bratter and Gorman 2011;
Campbell and Eggerling-Boeck 2006; Cheng and Lively 2009; Udry, Li, and
Hendrickson-Smith 2003).

Marginal man theorists viewed culture in monolithic and static terms and
thus to be trapped in between two cultures was seen as a difficult
experience. Contemporary views of culture instead emphasize the
flexible, fragmented, and fluid use of cultural bits that comprise a
``toolkit'' (DiMaggio 1997; Swidler 1986). From this perspective,
biracial individuals may actually benefit from an enlarged cultural
toolkit that allows them to navigate a variety of diverse social
contexts (Shih et al. 2019).

@@

Prior work on the outcomes for biracial individuals has been hampered by
several methodological issues. With good reason, researchers have been
reluctant to use data in which multiracial or biracial status is
determined solely by the self-identification of respondents. We know
that far fewer people identify as biracial than could feasibly do so
based on their parents' races (Morning and Saperstein 2018) and that
various selection effects operate on the decision to identify as
biracial among those who could (Davenport 2016; Roth 2005). Furthermore,
research on changes in racial reporting over time suggests that other
status markers like education and income may influence how people
identify themselves (Saperstein and Gullickson 2013; Saperstein and
Penner 2012). For example, a person with one Black and one White parent
who is routinely treated by others as Black and discriminated against as
Black may be more likely to identify themselves as exclusively Black
than a similar person of Black and White ancestry who experiences less
consistent discrimination. Thus, any analysis of outcomes for biracial
individuals based on self-identification must deal with the potential
for reverse causation in the observed differences between biracial
individuals and their constituent groups (Bratter 2018).

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
  letterpaper,
]{article}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

% -----------------------
% CUSTOM PREAMBLE STUFF
% -----------------------

\usepackage{orcidlink}
\definecolor{mypink}{RGB}{219, 48, 122}
\usepackage{endnotes}
\usepackage[nolists]{endfloat}

% ---------------------------
% END CUSTOM PREAMBLE STUFF
% ---------------------------
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[style=apa,]{biblatex}
\addbibresource{project.bib}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Differences in the Risk of Grade Retention for Biracial and Monoracial Students in the US, 2010-2019},
  pdfauthor={Aaron Gullickson},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Differences in the Risk of Grade Retention for Biracial and
Monoracial Students in the US, 2010-2019\thanks{Supplementary materials
provided with this article include full model results upon which figures
are based as well as any sensitivity analysis described herein. All code
and data for this project are available at
\url{https://osf.io/4fevt/?view_only=4abc6d86595c4313a8d4792471e9bc0d}.}}
\author{
Aaron Gullickson~\orcidlink{0000-0001-7237-8131}\\University of
Oregon, Sociology\\\href{mailto:aarong@uoregon.edu}{aarong@uoregon.edu}}
\date{}
\begin{document}

\maketitle
\begin{abstract}
Understanding how outcomes for biracial individuals compare to their
monoracial peers is critical for understanding how patterns of racial
inequality in the contemporary United States might be shifting. Yet, we
know very little about the life chances of biracial individuals due to
limitations in most available data sources. In this article, I utilize
American Community Survey data from 2010-2019 to examine the risk of
being clearly behind expected grade among biracial and monoracial K-12
students, helping to fill a gap in our understanding. With large sample
sizes for most biracial groups, I am able to estimate grade retention
risk for biracial students with enough precision to differentiate even
modest differences in risk relative to monoracial groups. The results
indicate that for most biracial groups, biracial students have risk
similar to their monoracial constituent group with lower risk. Although
biracial students tend to have favorable family resource
characteristics, controlling for these characteristics does little to
change the overall placement of their outcomes.
\end{abstract}

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi

\setstretch{1}
\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

The experiences of mixed race individuals will play an important role in
the future of race in the US. A rise in interracial unions since the
1960s has resulted in a ``biracial baby boom'' of individuals whose
parents cross racial lines \autocite{root_racially_1992} and this
demographic surge may have transformative effects on racial boundaries
\autocite{hochschild_creating_2012,alba_great_2020}. While racial mixing
is nothing new to US history
\autocite{morning_who_2000,gullickson_choosing_2011}, this biracial boom
is notable for its size and has occurred in an environment in which
historical norms of racial classification may be eroding.

The transformative potential of this biracial baby boom is contingent on
both how these biracial individuals identify their own race and how they
are sorted into a prevailing system of racial inequality. While
extensive scholarship documents the identification and classification
decisions made by or on behalf of multiracial individuals
\autocite{xie_racial_1997,qian_options_2004,roth_end_2005a,brunsma_interracial_2005,herman_forced_2004,holloway_place_2009,bratter_mother_2009,davenport_role_2016,liebler_boundaries_2016},
only a smattering of research has documented how the life chances of
biracial individuals compare to their monoracial peers
\autocite{kao_racial_1999,campbell_multiracial_2009,herman_blackwhiteother_2009}.
In this article, I expand on this work by examining the risk of grade
retention for K-12 students whose parents identify with different races.

By examining this outcome, I am able to overcome two important
limitations that have hampered prior work on the life chances of
biracial individuals. First, using self identification of biracial
respondents to analyze outcomes is problematic. Prior research indicates
that individuals whose parents belong to different races do not
consistently identify with both races
\autocite{doyle_are_2007,liebler_america_2017,gullickson_racial_2019}.
Furthermore, social status and social mobility can predict racial
identification generally
\autocite{saperstein_racial_2012,saperstein_mulatto_2013} and social
status is related to the identification choices of biracial individuals
specifically \autocite{roth_end_2005a,davenport_role_2016}. Thus,
differences in outcomes among biracial individuals may in part produce
differences in identification, rather than the other way around.

@@

Biracial children who reside with both biological parents can be
identified in large scale data sources such as the Census. Researchers
have previously utilized this feature to examine the classification
choices made on behalf of these children
\autocite{xie_racial_1997,roth_end_2005a}. Analyzing outcomes for such
children is more difficult, for the simple reason that they do not have
many measurable outcomes yet. However, grade retention is measurable and
important even for young children. Being held back in school is a
relatively common occurrence that has consequences for later educational
transitions \autocite{jimerson_metaanalysis_2001} and is
disproportionately applied by race \autocite{warren_patterns_2014}.
Grade retention is one of the earliest stratifiers of life chances that
an individual will experience.

@@

Discrimination and bias by race play an important role in overall levels
of racial inequality \autocite{quillian_new_2006,pager_sociology_2008}.
Discrimination need not be consciously recognized by actors, but may
instead reflect cognitive biases and schema that individuals unknowingly
rely upon and reinforce in social interactions
\autocite{quillian_new_2006}. Discrimination can also be systemic in
nature when biases are built into and reinforced by organizations and
institutional systems \autocite{reskin_race_2012}. In our current
``colorblind'' regime \autocite{bonilla-silva_racism_2006}, such systems
may be formally race neutral but produce disparate impacts by race
\autocite{pager_sociology_2008}.

@@

The strongest historical precedent for understanding such classification
in the US is the ``one-drop rule'' norm governing classification as
Black. According to the one-drop rule, individuals of known Black
ancestry are considered exclusively Black, regardless of other
ancestries \autocite{davis_who_1991}. \textcite{davis_who_1991}
generalizes this practice into the concept of \emph{hypodescent} in
which individuals with mixed racial ancestry will be classified as
members of the lower status group. A parallel pattern of
\emph{hyperdescent} implies that individuals of mixed racial ancestry
will be classified as members of the higher status group. This pattern
of hyperdescent has historically governed the experiences of individuals
with American Indian ancestry
\autocite{snipp_american_1989,wolfe_land_2001}.

Although the hypodescent/hyperdescent paradigm is often used to frame
the experiences of multiracial individuals, its generalizability and
contemporary relevance is questionable. Hypodescent and hyperdescent are
observed in two specific historical cases involving Black ancestry and
American Indian ancestry, respectively. As
\textcite{iverson_regimes_2022} note, these two cases could just as
parsimoniously be explained by a ``dominance'' model which ranks
ancestries by their tendencies to be supercessive or recessive in
determining identification and classification.

The development of particular norms may also reflect the historical
regime in which they were developed
\autocite{gullickson_choosing_2011,iverson_regimes_2022}. In this case,
neither the hypodescent/hyperdescent or dominance paradigm may help us
to fully understand the experiences of those of Latino and Asian
ancestry due to more recent processes of racialization. Furthermore, the
historical norms for all groups may be transforming in the post-Civil
Rights era.

@@

Second, biracial people's experience of discrimination may be
``softened'' because a more ambiguous physical appearance provokes less
racial antipathy. The US already has a long history of this kind of
softening. Because of the one-drop rule, Black individuals vary
substantially in skin tone and those individuals with a lighter skin
tone have better outcomes
\autocite{keith_skin_1991,hughes_significance_1990,monk_skin_2014} and
are viewed more favorably by both Blacks and Whites
\autocite{maddox_cognitive_2002a}. Thus, light skin moderates the stigma
of Blackness even among monoracial individuals.

Third, biracial individuals may be classified into an ambiguous
``middleman'' minority position. This kind of middle tier status for
individuals of mixed race is common in Latin America
\autocite{telles_race_2009} and \textcite{bonilla-silva_biracial_2004}
has argued that such a system may be emerging in the contemporary US as
well. In this case, many biracial groups may occupy a nebulous and
ill-defined middleman or buffer class position in-between whites and
darker-skinned non-whites, helping to cement a new kind of racial
hierarchy.

@@

Racial inequality is at least partly explained by different
distributions of material, cultural, and social resources across racial
groups \autocite{conley_being_1999}. On average, socioeconomic
background (e.g.~parental income, wealth, and education) differs by race
and such socioeconomic background is consequential for outcomes. Other
less materially defined resources also vary across groups such as
cultural styles, social networks, and neighborhood quality. These
differences in background resources reflect the accumulation of
historical processes of discrimination and, when combined with
contemporary processes of discrimination, magnify the overall level of
racial inequality we observe.

How do differences in the distribution of resources affect biracial
individuals? Naively, we might expect biracial individuals to inherit
parental resources roughly halfway between their two constituent groups,
on average \autocite{chew_american_1989}. Prior research looking at
biracial children in the Census data has found some evidence to support
this argument \autocite{chew_american_1989}. In such a case, the
resources of biracial individuals would predict outcomes in-between the
two constituent groups, barring other mechanisms.

@@

Status exchange theory predicts that given an unequal racial hierarchy,
individuals from a lower status racial group will be positively selected
on socioeconomic characteristics into intermarriage with a higher status
group, while individuals from a higher status racial group will be
similarly negatively selected into the same marriage
\autocite{davis_intermarriage_1941,merton_intermarriage_1941,fu_racial_2001}.
Prior research has shown substantial evidence of this pattern in the
case of Black male/White female marriages in the US, but weaker evidence
for other types of unions
\autocite{gullickson_education_2006,fu_racial_2001,kalmijn_educational_2010,hou_interracial_2011}.
Even in cases where status exchange is present, its net effect on
resources is complex because status exchange positively selects one
partner and negatively selects the other partner. Therefore the two
effects may somewhat cancel each other out in terms of the overall
distribution of resources across parents.

@@

The experience of hybridity itself may also positively or negatively
affect outcomes for biracial individuals. According to ``marginal man''
theory, biracial individuals will find it difficult to fit in due to the
dissonance of an identity that crosses strong and salient racial
boundaries \autocite{park_human_1928,stonequist_problem_1935}. This
liminal identity conflict will result in negative mental health
experiences that could feed into other negative outcomes as well.
Evidence for such negative mental health experiences, however, has been
mixed
\autocite{udry_health_2003,campbell_what_2006,cheng_multiracial_2009,bratter_does_2011}.

Marginal man theorists viewed culture in monolithic and static terms and
thus to be trapped in between two cultures was seen as a difficult
experience. Contemporary views of culture instead emphasize the
flexible, fragmented, and fluid use of cultural bits that comprise a
``toolkit'' \autocite{swidler_culture_1986,dimaggio_culture_1997}. From
this perspective, biracial individuals may actually benefit from an
enlarged cultural toolkit that allows them to navigate a variety of
diverse social contexts \autocite{shih_multiple_2019}.

@@

Prior work on the outcomes for biracial individuals has been hampered by
several methodological issues. With good reason, researchers have been
reluctant to use data in which multiracial or biracial status is
determined solely by the self-identification of respondents. We know
that far fewer people identify as biracial than could feasibly do so
based on their parents' races \autocite{morning_generational_2018} and
that various selection effects operate on the decision to identify as
biracial among those who could
\autocite{roth_end_2005a,davenport_role_2016}. Furthermore, research on
changes in racial reporting over time suggests that other status markers
like education and income may influence how people identify themselves
\autocite{saperstein_racial_2012,saperstein_mulatto_2013}. For example,
a person with one Black and one White parent who is routinely treated by
others as Black and discriminated against as Black may be more likely to
identify themselves as exclusively Black than a similar person of Black
and White ancestry who experiences less consistent discrimination. Thus,
any analysis of outcomes for biracial individuals based on
self-identification must deal with the potential for reverse causation
in the observed differences between biracial individuals and their
constituent groups \autocite{bratter_multiracial_2018}.
